---
title: "Penguin Visualization"
author: "Clemens S. Heithecker"
subtitle: "Module 1"
output:
  html_document:
    df_print: paged
---

# Setup

In this module we use the tidyverse package to manipulate data frames. We will
also use the palmerpenguins package which includes data about penguins.

```{r message = FALSE, warning = FALSE}
library(palmerpenguins)
library(tidyverse)
```

*Note*: You can install packages using the `install.packages("palmerpenguins")`
command.

# Data

We can now access the data by typing the command `data("penguins")`. In the top-
right area of the RStudio screen under the tap "Environment" we can see two data
sets.

- *penguins* with 344 observations and 8 variables
- *penguins_raw* with 344 observations and 17 variables

We will only use the first data set (penguins). Take a look at the data set by
double clicking on the `penguins` data set in the environment or by using the
following commands:

- `View(penguins)` which opens the data frame
- `head(penguins)` which outputs the top 5 rows of the data set
- `tail(penguins)` which outputs the bottom 5 rows of the data set
- `summary(penguins)` which ouputs a table of summary statistics

```{r}
data("penguins")

head(penguins)
tail(penguins)
```

```{r}
summary(penguins)
```

The data set contains 8 variables and 344 individual penguins. If you want to
know what all variables mean you can use the command `?penguins` to inspect the
documentation.

```{r}
?penguins
```

There are some `NA` or missing values in the data set. We can remove those
observations using the following code:

```{r}
penguins <- penguins %>%
  na.omit()
```

Now the number of observations decreased to 333 meaning we lost 11 penguins.

# Plotting Data

In 2005  Hadley Wickham created ggplot as part of his PhD. `ggplot` is a part of
the tidyverse package and allows you to plot anything you can imagine.

Any plot in ggplot has (at least) the following parts:

- the command `ggplot()` which tells R to plot.
- a `geom_` that tells ggplot which type of graph should be created. Examples are

  - `geom_point()` for a scatter plot
  - `geom_line()` for a line graph
  - `geom_bar()` for a bar plot
  - `geom_histogram()` for a histogram
  - `geom_boxplot()`

We need to specify how to plot the variables in our graph. In ggplot we can link
an **aesthetic** of a graph to a column in a data frame. Examples include

- x-coordinate
- y-coordinate
- size and shape of points
- fill color
- line color
- line size
- transparency (alpha value)
- line type

You can specify the aesthetic within the `ggplot()` function call or the
`geom_`. It is often easier to use the aesthetics argument within the `geom_`
call especially when dealing with multiple `geom_` in one plot.

## Scatter Plot

Suppose we are interested in the correlation between flipper length and bill
length. We assume that animals with a larger flipper will also have a larger
bill.

We can plot the data using a scatter plot `geom_scatter()`.

```{r}
ggplot(data = penguins) +
  geom_point(mapping = aes(x = flipper_length_mm, y = bill_length_mm))
```

## Adding Color to the Plot

We could experiment with adding colors and shapes to this plot. Let's make the
points of the scatter plot red, 3 times bigger and change the shape.

```{r}
ggplot(data = penguins) +
  geom_point(mapping = aes(x = flipper_length_mm, y = bill_length_mm),
             color = "red", size = 3, shape = 1)
```

Because we added the color, size and shape attributes outside the `aes()`
function we apply the same attributes to all data points.

## Adding Dimensions of Species and Sex

It seems that there is a positive correlation between flipper and bill length.
Let's examine if there are group differences between sex and species. In our
next plot we will add an aesthetic `color` for species and `shape` for sex to
our plot. We should do this inside the `aes()` function because we now want to
make the color and the shape dependent on the variables `species` and `sex`.

```{r}
ggplot(data = penguins,
       mapping = aes(x = flipper_length_mm, y = bill_length_mm)) +
  geom_point(mapping = aes(color = species, shape = sex),
             size = 3)
```

## Improving the Plot

We can make the plot nicer by adding labels with `labs()`. We can add

- a title
- a subtitle
- an x-axis label
- a y-axis label
- a caption

We can also add a theme. Check out some examples at
[https://ggplot2.tidyverse.org/reference/ggtheme.html](https://ggplot2.tidyverse.org/reference/ggtheme.html).

```{r}
ggplot(data = penguins,
       mapping = aes(x = flipper_length_mm, y = bill_length_mm)) +
  geom_point(size = 3,
             mapping = aes(color = species, shape = sex)) +
  labs(
    title = "Bill length vs flipper length",
    subtitle = "Clemens S. Heithecker",
    x = "Flipper length (mm)",
    y = "Bill length (mm)",
    caption = "Source: Penguins"
  ) +
  theme_bw()
```

# Box Plots

Let's make a box plot.

```{r}
ggplot(data = penguins, mapping = aes(x = species, y = body_mass_g)) +
  geom_boxplot(mapping = aes(fill = sex)) +
  labs(
    x = "Species",
    y = "Body mass in grams"
  )
```

We could also add another dimension by making a plot per island. With this plot
we are able to see which species lives on which island and whether species
living on different islands have different body masses. We can split the plots
per island by using the `facet_wrap(~ island)`.

```{r}
ggplot(data = penguins, mapping = aes(x = species, y = body_mass_g)) +
  geom_boxplot(mapping = aes(fill = sex)) +
  facet_wrap(~ island) +
  labs(
    x = "Species",
    y = "Body mass in grams"
  )
```

Let's change the colors of the `sex` variable.

```{r}
ggplot(data = penguins, mapping = aes(x = species, y = body_mass_g)) +
  geom_boxplot(mapping = aes(fill = sex)) +
  facet_wrap(~ island) +
  scale_fill_manual(values = c("mediumblue", "#DA020E")) +
  labs(
    x = "Species",
    y = "Body mass in grams"
  )
```










