---
title: "Reading Data"
author: "Clemens S. Heithecker"
subtitle: "Module 1"
output:
  html_document:
    df_print: paged
---

# Setup

In this module we use the tidyverse package to manipulate data frames. We will
also use the readxl package to read Excel files.

```{r message = FALSE, warning = FALSE}
library(readxl)
library(tidyverse)
```

# Read CSV files

Let's start off by reading the content of the **CSV* file `people.csv`. This
file contains a list of people and their corresponding age.

Use the command `data_frame <- read.csv("path/file_name.csv")`, where
`data_frame` is the name you want to call the resulting data frame and
`path/file_name.csv` the file name including the path of the CSV file you want
to read in.

*Note*: Some CSV files do not use a comma as a delimiter. For example, some
European regions use a semicolon (";") to separate values. In this case, use the
function `read.csv2()` instead of `read.csv()`.

# Exercise 1—Reading a CSV File

Read the file `people.csv` from the data folder and call the data frame
"people".

```{r}
people <- read.csv("data/people.csv")
```

By typing `people` we can view the data. Alternatively, you can use the
`print()` function.

```{r}
people
```

To get an impression of the data we can either use the `head(data_frame)` or the
`tail(data_frame)` function. These functions return the first or last few rows
of a given data frame.

```{r}
head(people)
```

```{r}
tail(people)
```

The function `read.csv2()` allows for an optional argument to specify the
separator or delimiter.

```{r}
people <- read.csv2("data/people.csv", sep = ",")

people
```

# Read Excel Files

An **Excel** file is another common data file format. We can load Excel files
using the `read_excel()` function from the `readxl` package. We already imported
the package at the start of the document.

Let's read the `colors.xlsx` file and assign it to a new data frame called
`colors`. Then, we can use the `head()` function to check out the first 5 rows
of the new data frame.

```{r}
colors <- read_excel("data/colors.xlsx")

colors
```

Looking at the data, we see that the Excel sheet is read into a new data frame
including empty rows. we can repair this using the package `dplyr` which is part
of the `tidyverse` package but we can also use the `skip` argument inside the
`read_excel` function. In this case, we do not want to read the first 4 rows,
so we should add `skip = 4` to our function call.

```{r}
colors <- read_excel("data/colors.xlsx", skip = 4)

head(colors)
```

## Reading Separate Sheets

If you have opened the Excel sheet, you might have noticed that it has two
worksheets. We only read the first sheet, but we can also read the second
worksheet.

```{r}
colors_light <- read_excel("data/colors.xlsx", sheet = "400", skip = 4)

head(colors_light)
```

*Note*: Instead of `sheet = "400"` we could have used `sheet = 2` to select the
second sheet of the Excel file.

# Writing Data to a File

It is also possible to write a data frame as a CSV file by using the command
`write.csv()` (or `write.csv2()`). This is not often needed becuase all the data
editing is reproducible within the script. But sometimes you may want to write
your data as a CSV file and you can do that like the following example:

```{r}
write.csv(colors, file = "data/colors.csv")
```

# Data Structures

In R there are only a few common types of data:

- `character` (string)
- `numeric` (real numbers)
- `integer` (integer numbers)
- `complex` (complex numbers)
- `logical` (binary `TRUE` or `FALSE` or `NA`)
- `factor` (categorical data, e.g. *male* and *female*)

Using the command `str()` you can get the structure of a given data frame.

Let's look at an example. First, read in the `gender.csv` file and assign it to
a new data frame called `gender`.

```{r}
gender <- read.csv("data/gender.csv")

gender
```

Now check the structure of the data frame `gender`.

```{r}
str(gender)
```

We can see that the `gender` column within the `gender` data frame has a data
type of *character* and the `number` column is of type *intege*. 

## Changing the Structure of a Data Frame

We can change the data structure of variables in several ways.

Suppose we want to change the variable `gender` into a string of characters.

```{r}
gender$gender <- as.character(gender$gender)
```

And now suppose we want to change the `number` column into a factor.

```{r}
gender$number <- as.factor(gender$number)
```

Let's call the `str()` function again to see how the data structures changed.

```{r}
str(gender)
```

*Important note*: To change a factor variable to a numeric variable, we first
need to convert it into a character variable.

```{r}
gender$number2 <- as.numeric(as.character(gender$number))

gender
```

# Data Summary

We can use the `summary()` function to calculate summary statistics for each
variable or column of a data frame. For example:

```{r}
summary(people)
```

# Read RDS Files

The RDS file is a special data format specific to R. You can load in data from a
RDS file using the `readRDS` function.

```{r}
people_rds <- readRDS("data/people.rds")

summary(people_rds)
```




